A fully labelled proof system for intuitionistic modal logics

### 2021-01-02 FullyLabelled/JLC/introduction.tex
working on review

However,
in this paper we are concerned with \emph{intuitionistic modal
logics} whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in 
Kripke frames for modal logics, and a relation $\le$ that
corresponds to the preorder relation in Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
	\item The transitivity rule in~\eqref{eq:Rtrans} can be axiomatised by the conjunction of the two
	versions of the $\fourax$-axiom
	\begin{equation}
	\label{eq:4ax}
	\ax{4^{\BOX}}\colon\fm{\BOX A\IMP \BOX\BOX A}
	\qquand
	\rn{4^{\DIA}}\colon\fm{\DIA\DIA A\IMP\DIA A}
%	\quadcm
	\end{equation}
	which are equivalent in classical
	modal logic. However, in intuitionistic modal logic they are not
	equivalent, and even though the logic $\IKfour$ 
	(intuitionistic version of modal logic $\Kfour$) contains both
	axioms, they can also be added independently to the logic $\IK$ 
	(intuitionistic version of $\K$). The proof theory of these
	distinct logics has not been discussed before, but their
	corresponding frame conditions
	\begin{equation}
	\label{eq:4frame}
	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
	\quand
	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
	\;
	\end{equation}
	%  \todo{add graphical representation of the conditions}
	$$
	\xymatrix{
		x' \ar@{.>}[drr]^R \\
		x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
	}
	\hskip10em
	\xymatrix{
		&& z''  \\
		x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
	}
	$$
	respectively, have been studied in~\cite{plotkin:stirling:86}.
	%  \sonia{the general condition is correct but not the application to 4}
	\item The correspondence between the syntax and the semantics is not
	as clean as one would expect. As only the $R$-relation (and not the
	$\le$-relation) is visible in an ordinary labelled
	sequent, we only have the result that a formula is provable in an extension of $\IK$ beyond paths axioms 
	if and only if is satisfied in all \emph{graph-consistent}\footnote{This
		means that every layer in the model can be lifted to any future of
		any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
	models satisfying the corresponding conditions, 
	%  \sonia{this is only true for logics outside of the path axioms.}
	%  \todo{check that Simpson's system is sound and complete for all logics defined by path axioms. Is it that the axiomatisation is also sound and complete for every models and not only graph-consistent ones?}
	%  \sonia{yes, they are sound and complete for the general models, but they do not \emph{characterize} the frames, they are not \emph{frame conditions}, as opposed to the \emph{correspondence result} of Plotkin and Sterling.}
	as observed by Simpson
	and considered an inelegant solution~\cite{simpson:phd}. % (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two concerns we propose here to enrich usual 
labelled sequents by allowing both, relational atoms of the
form $\futs xy$ and of the form $\accs xy$. 
Consequently, following the same ``axiom-to-rule'' method as Simpson, we can translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
\label{eq:4rules}
\vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
\qquand
\vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that had not been treated proof-theoretically before.  
Furthermore, the relation
between syntax and semantics can be recovered as one would expect: A formula is
provable in this system if and only if it is valid in all models satisfying the corresponding frame condition.

>>>>

However,
in this paper we are concerned with \emph{intuitionistic modal
logics} whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in 
Kripke frames for modal logics, and a relation $\le$ that
corresponds to the preorder relation in Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
	\item The transitivity rule in~\eqref{eq:Rtrans} can be axiomatised by the conjunction of the two
	versions of the $\fourax$-axiom
	\begin{equation}
	\label{eq:4ax}
	\ax{4^{\BOX}}\colon\fm{\BOX A\IMP \BOX\BOX A}
	\qquand
	\rn{4^{\DIA}}\colon\fm{\DIA\DIA A\IMP\DIA A}
%	\quadcm
	\end{equation}
	which are equivalent in classical
	modal logic. However, in intuitionistic modal logic they are not
	equivalent, and even though the logic $\IKfour$ 
	(intuitionistic version of modal logic $\Kfour$) contains both
	axioms, they can also be added independently to the logic $\IK$ 
	(intuitionistic version of $\K$). The proof theory of these
	distinct logics has not been discussed before, but their
	corresponding frame conditions
	\begin{equation}
	\label{eq:4frame}
	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
	\quand
	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
	\;
	\end{equation}
	%  \todo{add graphical representation of the conditions}
	$$
	\xymatrix{
		x' \ar@{.>}[drr]^R \\
		x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
	}
	\hskip10em
	\xymatrix{
		&& z''  \\
		x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
	}
	$$
	respectively, have been studied in~\cite{plotkin:stirling:86}.
	%  \sonia{the general condition is correct but not the application to 4}
	\item The correspondence between the syntax and the semantics is not
	as clean as one would expect. As only the $R$-relation (and not the
	$\le$-relation) is visible in an ordinary labelled
	sequent, we only have the result that a formula is provable in an extension of $\IK$ beyond paths axioms 
	if and only if is satisfied in all \emph{graph-consistent}\footnote{This
		means that every layer in the model can be lifted to any future of
		any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
	models satisfying the corresponding conditions, 
	%  \sonia{this is only true for logics outside of the path axioms.}
	%  \todo{check that Simpson's system is sound and complete for all logics defined by path axioms. Is it that the axiomatisation is also sound and complete for every models and not only graph-consistent ones?}
	%  \sonia{yes, they are sound and complete for the general models, but they do not \emph{characterize} the frames, they are not \emph{frame conditions}, as opposed to the \emph{correspondence result} of Plotkin and Sterling.}
	as observed by Simpson
	and considered an inelegant solution~\cite{simpson:phd}. % (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two concerns we propose here to enrich usual 
labelled sequents by allowing both, relational atoms of the
form $\futs xy$ and of the form $\accs xy$.

\marianela{The reviewer suggested to change the above sentence for the next one:
	\emph{"As done earlier in the literature  on intuitionistic modal logic (cf....) we consider here
	to enrich"} since Maffezioli, Naibo and Negri have already done this before and the next one:

	D. Garg, V. Genovese, and S. Negri, Countermodels from sequent calculi in multi-modal	logics. LICS 2012, IEEE Computer Society, pp. 315–324, 2012. (ALREADY ADDED. SHOULD I REMOVE IT?)
	What do you think about changing the sentence? We have already mentioned Maffezioli, Naibo and Negri's paper a few times.
	}

%%
Consequently, following the same ``axiom-to-rule'' method as Simpson, we can translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
\label{eq:4rules}
\vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
\qquand
\vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that had not been treated proof-theoretically before.  
%%
Furthermore, the relation
between syntax and semantics can be recovered as one would expect: A formula is
provable in this system if and only if it is valid in all models satisfying the corresponding frame condition.
